\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-11-17 submission/manuscript.Rmd
Summary shortened.

## Summary
Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited by diverse microbial communities. Most studies focusing on epiphytic communities of macrophytes did not take into account temporal changes or applied low sampling frequency approaches. Illumina sequencing of the V4 16S rRNA region was performed to determine the seasonal dynamics of epiphytic communities of the seagrass *Cymodocea nodosa* and invasive macroalga *Caulerpa cylindracea*. Leaves and thalli were sampled in a meadow of *Cymodocea nodosa* invaded by the invasive *Caulerpa cylindracea* and in a monospecific settlement of *Caulerpa cylindracea* located in the proximity of the meadow at monthly intervals. For comparison the ambient prokaryotic plankton community was also characterized. At the OTU level, the microbial community composition differed between the ambient water and the epiphytic communities exhibiting host-specificity. Also, successional changes were observed connected to the macrophyte growth cycle. Taxonomic analysis, however, showed similar high rank groups in the ambient water and the epiphytic communities, with the exception of *Desulfobacterota*, which were only found on *Caulerpa cylindracea*. *Cyanobacteria* showed seasonal changes while other high rank taxa were present throughout the year. Phylogenetic groups present throughout the year constituted most of the sequences, while less abundant taxa showed seasonal patterns connected to the macrophyte growth cycle. Taken together, epiphytic microbial communities of the seagrass *Cymodocea nodosa* and the macroalgae *Caulerpa cylindracea* appear to be host-specific and contain taxa that undergo successional changes.

>>>>

## Summary
Surfaces of marine macrophytes are inhabited by diverse microbial communities. Most studies focusing on epiphytic communities of macrophytes did not take into account temporal changes or applied low sampling frequency approaches. The seasonal dynamics of epiphytic microbial communities was determined in a meadow of *Cymodocea nodosa* invaded by *Caulerpa cylindracea* and in a monospecific settlement of *Caulerpa cylindracea* at monthly intervals. For comparison the ambient prokaryotic picoplankton community was also characterized. At the OTU level, the microbial community composition differed between the ambient water and the epiphytic communities exhibiting host-specificity. Also, successional changes were observed connected to the macrophyte growth cycle. Taxonomic analysis, however, showed similar high rank groups in the ambient water and the epiphytic communities, with the exception of *Desulfobacterota*, which were only found on *Caulerpa cylindracea*. *Cyanobacteria* showed seasonal changes while other high rank taxa were present throughout the year. Phylogenetic groups present throughout the year constituted most of the sequences, while less abundant taxa showed seasonal patterns connected to the macrophyte growth cycle. Taken together, epiphytic microbial communities of the seagrass *Cymodocea nodosa* and the macroalga *Caulerpa cylindracea* appear to be host-specific and contain taxa that undergo successional changes.

### 2020-11-17 submission/manuscript.tex
Summary shortened.

%\usepackage{lineno}
%\renewcommand{\linenumberfont}{\normalfont\tiny}

@@

Surfaces of marine macrophytes (seagrasses and macroalgae) are inhabited
by diverse microbial communities. Most studies focusing on epiphytic
communities of macrophytes did not take into account temporal changes or
applied low sampling frequency approaches. Illumina sequencing of the V4
16S rRNA region was performed to determine the seasonal dynamics of
epiphytic communities of the seagrass \emph{Cymodocea nodosa} and
invasive macroalga \emph{Caulerpa cylindracea}. Leaves and thalli were
sampled in a meadow of \emph{Cymodocea nodosa} invaded by the invasive
\emph{Caulerpa cylindracea} and in a monospecific settlement of
\emph{Caulerpa cylindracea} located in the proximity of the meadow at
monthly intervals. For comparison the ambient prokaryotic plankton
community was also characterized. At the OTU level, the microbial
community composition differed between the ambient water and the
epiphytic communities exhibiting host-specificity. Also, successional
changes were observed connected to the macrophyte growth cycle.
Taxonomic analysis, however, showed similar high rank groups in the
ambient water and the epiphytic communities, with the exception of
\emph{Desulfobacterota}, which were only found on \emph{Caulerpa
cylindracea}. \emph{Cyanobacteria} showed seasonal changes while other
high rank taxa were present throughout the year. Phylogenetic groups
present throughout the year constituted most of the sequences, while
less abundant taxa showed seasonal patterns connected to the macrophyte
growth cycle. Taken together, epiphytic microbial communities of the
seagrass \emph{Cymodocea nodosa} and the macroalgae \emph{Caulerpa
cylindracea} appear to be host-specific and contain taxa that undergo
successional changes.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 17,082
different OTUs with 379 to 2,017 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,698.7 ±
139.6 OTUs{]} and monospecific {[}Number of OTUs, 1,735.2 ± 160.1
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,052.8 ± 216.1
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 528.8 ± 145.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (37.8 -- 53.6 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.54, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 53.6
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.

>>>>

\usepackage{lineno}
\renewcommand{\linenumberfont}{\normalfont\tiny}

@@

Surfaces of marine macrophytes are inhabited by diverse microbial
communities. Most studies focusing on epiphytic communities of
macrophytes did not take into account temporal changes or applied low
sampling frequency approaches. The seasonal dynamics of epiphytic
microbial communities was determined in a meadow of \emph{Cymodocea
nodosa} invaded by \emph{Caulerpa cylindracea} and in a monospecific
settlement of \emph{Caulerpa cylindracea} at monthly intervals. For
comparison the ambient prokaryotic picoplankton community was also
characterized. At the OTU level, the microbial community composition
differed between the ambient water and the epiphytic communities
exhibiting host-specificity. Also, successional changes were observed
connected to the macrophyte growth cycle. Taxonomic analysis, however,
showed similar high rank groups in the ambient water and the epiphytic
communities, with the exception of \emph{Desulfobacterota}, which were
only found on \emph{Caulerpa cylindracea}. \emph{Cyanobacteria} showed
seasonal changes while other high rank taxa were present throughout the
year. Phylogenetic groups present throughout the year constituted most
of the sequences, while less abundant taxa showed seasonal patterns
connected to the macrophyte growth cycle. Taken together, epiphytic
microbial communities of the seagrass \emph{Cymodocea nodosa} and the
macroalga \emph{Caulerpa cylindracea} appear to be host-specific and
contain taxa that undergo successional changes.

@@

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of sequences without
known relatives (no relative sequences) and eukaryotic, chloroplast and
mitochondrial sequences (\autoref{supp-nseq_notus}). A total of 35
samples originating from epiphytic archaeal and bacterial communities
associated with surfaces of the seagrass \emph{C. nodosa} and the
macroalga \emph{C. cylindracea} were analyzed. In addition, 18 samples
(one of the samples was sequenced twice) originating from the ambient
seawater were also processed for comparison. The number of reads per
sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off as commonly observed in
high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences as mentioned above reads were clustered into 28,750
different OTUs at a similarity level of 97 \si{\percent}. Read numbers
were normalized to the minimum number of sequences (8,408,
\autoref{supp-nseq_notus}) through rarefaction resulting in 16,986
different OTUs with 361 to 2,026 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes in richness
and diversity the observed number of OTUs, Chao1, ACE, Exponential
Shannon and Inverse Simpson (Jost, 2006) were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,678.1 ±
143.6 OTUs{]} and monospecific {[}Number of OTUs, 1,734.2 ± 150.9
OTUs{]}) than for \emph{C. nodosa} (Number of OTUs, 1,054.2 ± 213.0
OTUs) and lowest values were obtained for the microbial community of the
ambient seawater (Number of OTUs, 529.1 ± 136.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not reveal such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly higher values
in spring and summer than in autumn and winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient, respectively, were calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from the
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between ambient seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in-between the values of mixed
and monospecific \emph{C. cylindracea}. To assess seasonal changes in
the proportion of shared OTUs and communities the Jaccard's and
Bray-Curtis Similarity Coefficients were calculated between consecutive
sampling points (\autoref{shared}). Both coefficients showed similar
trends. Temporal proportional changes were more pronounced for ambient
seawater than for \emph{C. nodosa} and especially \emph{C. cylindracea}
associated communities (\autoref{shared}). In addition, only 0.4 -- 1.0
\si{\percent} of OTUs from each surface associated community were
present at all seasons. These persistent OTUs constituted a high
proportion of total sequences (36.9 -- 53.0 \si{\percent}). To further
disentangle the environmental and seasonal community dissimilarity a
Principal Coordinates Analysis (PCoA) based on Bray-Curtis distances and
OTU abundances was applied. A clear separation between ambient seawater
and surface associated communities was found (\autoref{pcoa}). In
addition, a separation of epiphytic bacterial and archaeal communities
based on host species was detected. This separation was further
supported by ANOSIM (R = 0.96, \emph{p} \textless{} 0.001). Seasonal
changes of \emph{C. nodosa} associated communities indicated a
separation between spring, summer and autumn/winter samples (ANOSIM, R =
0.53, \emph{p} \textless{} 0.01). For \emph{C. cylindracea} associated
communities a separation between summer and autumn/winter/spring samples
was observed that was, however, not as strong as for \emph{C. nodosa}
associated communities (ANOSIM, R = 0.31, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal changes in richness in the epiphytic community were substantial
as indicated by the proportion of OTUs (\(\leq\) 1.0 \si{\percent})
present at every sampling date. These persistent OTUs, however, were
accounting for a high proportion of sequences (\(\leq\) 53.0
\si{\percent}) (\autoref{shared}). A very similar proportion of
persistent OTUs was reported in high-frequency sampling studies
describing seasonal changes in picoplankton (Gilbert \emph{et al.},
2009, 2012). In comparison to the seawater community, a lower degree of
seasonal shifts was observed for the macrophyte surface associated
communities. It appears that microniches at the surfaces of macrophytes
are providing more stable conditions than the ambient seawater. At the
level of OTUs seasonal changes of \emph{C. nodosa} and \emph{C.
cylindracea} associated communities were identified that could be linked
to the growth cycle of the seagrass and macroalgae (M. Najdek, personal
communication). \emph{C. nodosa} was characterized by a spring community
during maximum seagrass proliferation, a summer community during the
highest standing stock of \emph{C.nodosa} and an autumn/winter community
during the decay of seagrass biomass. In contrast, \emph{C. cylindracea}
started to proliferate in late spring and was characterized only by a
summer community during high growth rates and by an autumn/winter/spring
community when the biomass was at the peak and decaying thereafter.
Similar seasonal changes in the epiphytic community were also described
for other macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.},
2011). Higher seasonal stability of \emph{C. cylindracea} surface
communities than in \emph{C. nodosa} was also observed in the higher
proportion of shared communities between two consecutive sampling dates
in \emph{C. cylindracea}.
